Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict

### 2020-09-22 paper.Rmd
drop in references

B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkably accurately". Indeed, B&S argue that this theoretical model outperforms four other "more mechanical" alternatives.They also claim that the integration of structure with process is better than alternatives over short forecasting windows. Third, they preregistered the list of thirty countries that have the highest risk of civil war onset. They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects.\footnote{\cite{PITF},\cite{ICEWS},\cite{VIEWS}\cite{others}} B\&S claim to be unique in assessing these forecasts. A qualitative analysis of their predictions allows them to conclude that their model is robust. We will return to their analysis later, after correcting the procedural mistakes we found in their research process.

@@

B\&S advocate the use of theory to guide prediction. But "theory" is not defined by B\&S and may be an ambiguous and undefined concept. `Theory'' has many interpretations in social science.  For some---the EITM movement, for example [@aldrich:etal:2008]---theory is a system of internally consistent, logically connected propositions from which additional statements may be deduced.  The additional statements can be subjected to empirical scrutiny.  An example might be William Riker's theory of coalitions [@riker:1962] which posits propositions in game theoretic form from which the size principal of minimal winning coalitions may be deduced and further examined empirically. A more recent example of strong theory is found in *The Political Logic of Survival* [@bdm:etal:2005]. For the purposes of discussion we call this strong theory.

@@

At this level of detail the theoretical model is more about choosing the correct variables and this is the contrast between theoretical prediction models and atheoretical machine learning that B&S focus on. But this dichotomy is misguided. The data that are used in a predictive model have to come from somewhere. Even if the inclusion of some factors is not explicitly justified, it is hardly the case that one assembles data without the expectation that it will be useful. It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well. Rather, we posit that most of the cases B&S might characterize as atheoretical enter the "theoretical" , with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}

Another way in which we could contrast theoretical from atheoretical prediction models is in terms of their flexibility in regard to functional form. Theory should not only tell us what factors matter, but also how they relate to each other. We think that few would be strongly committed to the idea that the linear additive equations, maybe with the occasional polynomial or multiplicative interaction terms, that dominate published regression analyses, are *the* correct functional specification. Is the danger posed by low-level violence the same when tensions are high, with high demands and high non-violent repression, as when tensions are low? Are those relationships linear in the log-odds? Some machine learning models, random forests included, are flexible in this regard and to some extent capable of learning non-linear and interactive relationships. B&S implicitly concede on this point since they use random forests for both their Escalation model and the atheoretical alternatives.

>>>>

B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkably accurately". Indeed, B&S argue that this theoretical model outperforms four other "more mechanical" alternatives.They also claim that the integration of structure with process is better than alternatives over short forecasting windows. Third, they preregistered the list of thirty countries that have the highest risk of civil war onset. They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects.^[PITF [e.g. @goldstone:etal:2010; @beger:ward:2017], the W-ICEWS project [@ward:etal:2013], VIEWS [@hegre:etal:2019], and others, e.g. @beger:etal:2020.] B\&S claim to be unique in assessing these forecasts. A qualitative analysis of their predictions allows them to conclude that their model is robust. We will return to their analysis later, after correcting the procedural mistakes we found in their research process.

@@

B\&S advocate the use of theory to guide prediction. But "theory" is not defined by B\&S and may be an ambiguous and undefined concept. "Theory" has many interpretations in social science.  For some---the EITM movement, for example [@aldrich:etal:2008]---theory is a system of internally consistent, logically connected propositions from which additional statements may be deduced.  The additional statements can be subjected to empirical scrutiny.  An example might be William Riker's theory of coalitions [@riker:1962] which posits propositions in game theoretic form from which the size principal of minimal winning coalitions may be deduced and further examined empirically. A more recent example of strong theory is found in *The Political Logic of Survival* [@bdm:etal:2005]. For the purposes of discussion we call this strong theory.

@@

At this level of detail the theoretical model is more about choosing the correct variables and this is the contrast between theoretical prediction models and atheoretical machine learning that B&S focus on. But this dichotomy is misguided. The data that are used in a predictive model have to come from somewhere. Even if the inclusion of some factors is not explicitly justified, it is hardly the case that one assembles data without the expectation that it will be useful. It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well. Rather, we posit that most of the cases B&S might characterize as atheoretical enter the "theoretical", with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}

Another way in which we could contrast theoretical from atheoretical prediction models is in terms of their flexibility in regard to functional form. Theory should not only tell us what factors matter, but also how they relate to each other. We think that few would be strongly committed to the idea that the linear additive equations that dominate published regression analyses, maybe with the occasional polynomial or multiplicative interaction terms, are *the* correct functional specification. Is the danger posed by low-level violence the same when tensions are high, with high demands and high non-violent repression, as when tensions are low? Are those relationships linear in the log-odds? Some machine learning models, random forests included, are flexible in this regard and to some extent capable of learning non-linear and interactive relationships. B&S implicitly concede on this point since they use random forests for both their Escalation model and the atheoretical alternatives. But in the absence of strong a priori theoretical expectations about functional form, why not use machine learning tools to try to uncover any such non-linear relationships in need of explanation, and use this to improve our theory?
